Learning \strips\ action models from {\em state-invariants}

### 2019-02-08 doc/ijcai-19-deductivelearning/deductivelearning.tex
first full version (no experiments though) of the ijcai19 paper on deductive learning

\begin{abstract}
This paper addresses the learning of action models from {\em state-invariants} (i.e logic formulae that specify constraints about the possible states of a given domain) to cushion the negative impact of insufficient learning examples. Our approach is a {\em classical planning} compilation that is flexible to different kinds of input knowledge (e.g., partially observations of plan executions including partially observed intermediate states and/or actions) and outputs an action model that is {\em consistent} with the given input knowledge. The experimental results show that, even at unfavorable scenarios where input observations are minimal (just an {\em initial state} and the {\em goals}), {\em state-invariant} are helpful to learn better \strips\ action models.
\end{abstract}

@@

The specification of planning action models is a complex process that limits, too often, the application of {\em model-based planning} systems to real-world tasks~\cite{kambhampati:modellite:AAAI2007}. The {\em machine learning} of action models can relieve the {\em knowledge acquisition bottleneck} of planning and nowadays, there exists a wide range of effective approaches for learning action models~\cite{arora:amodels:ker2018}. Many of the most successful approaches for learning planning action models are however purely {\em inductive}~\cite{yang2007learning,pasula2007learning,mourao2010learning,zhuo2013action}, meaning that their performance is linked to the {\em amount} and {\em quality} of the input examples (which normally are observations of plan executions generated by the aimed action model).

This paper addresses the learning of action models exploiting a different source of knowledge, {\em deductive} knowledge, with the form of {\em state-invariants} (i.e. logic formulae that specify constraints about the possible states of a given domain) to cushion the negative impact of insufficient learning examples. Given an action model, state-of-the-art planners infer {\em state-invariants} from that model to reduce search spaces and make the planning process more efficient~\cite{helmert2009concise}. In this paper we follow the opposite direction and leverage {\em state-invariants} to learn the planning action model.

Our approach for learning \strips\ action models from {\em state-invariants} is building a classical planning compilation that is inspired by recent work by~\citeauthor{aineto2018learning}~\citeyear{aineto2018learning}. Our compilation is flexible to different kinds of input knowledge (e.g., partially/fully observations of actions of plan executions as well as partially/fully observed intermediate states) and outputs an action model that is {\em consistent} with the given input knowledge. The experimental results show that, even at unfavorable scenarios where input observations are minimal (just an {\em initial state} and the {\em goals}), {\em state-invariant} help to learn better \strips\ models with the {\em classical planning} compilation.

@@

{\em State invariants} is a kind of state-constraints useful for computing more compact state representations~\cite{helmert2009concise} or making {\em satisfiability planning} and {\em backward search} more efficient~\cite{rintanen2014madagascar,alcazar2015reminder}. Given a classical planning problem $P=\tup{F,A,I,G}$, a {\em state invariant} is a formula $\phi$ that holds at the initial state of a given classical planning problem, $I\models \phi$, and at every state $s$, built from $F$, that is reachable from $I$ by applying actions in $A$.

The formula $\phi_{I,A}^*$ represents the {\em strongest invariant} and exactly characterizes the set of all states reachable from $I$ with the actions in $A$. For instance Figure~\ref{fig:strongest-invariant} shows five clauses that define the {\em strongest invariant} for the {\em blocksworld} planning domain~\cite{slaney2001blocks}. There are infinitely many strongest invariants, but they are all logically equivalent, and computing the strongest invariant is PSPACE-hard (as hard as testing plan existence~\cite{bylander:complexity:AIJ1994}).

\begin{figure}[hbt!]
  \begin{footnotesize}
$\forall x_1,x_2\ ontable(x_1)\leftrightarrow\neg on(x_1,x_2)$.\\
$\forall x_1,x_2\ clear(x_1)\leftrightarrow\neg on(x_2,x_1)$.\\
$\forall x_1,x_2,x_3\ \neg on(x_1,x_2)\vee\neg on(x_1,x_3)\ such\ that\ x_2\neq x_3$.\\
$\forall x_1,x_2,x_3\ \neg on(x_2,x_1)\vee\neg on(x_3,x_1)\ such\ that\ x_2\neq x_3$.\\
$\forall x_1,\ldots,x_n\ \neg(on(x_1,x_2)\wedge on(x_2,x_3)\wedge\ldots\wedge on(x_{n-1},x_n)\wedge on(x_n,x_1)).$
\end{footnotesize}
 \caption{\small {\em Strongest invariant} for the {\em blocksworld} domain.}
\label{fig:strongest-invariant}
\end{figure}

A {\em mutex} (mutually exclusive) is a state invariant that takes the form of a binary clause and indicates a pair of different properties that cannot be simultaneously true~\cite{kautz:mutex:IJCAI1999}. For instance in a three-block {\em blocksworld}, $\phi_1=\neg on(block_A,block_B)\vee \neg on(block_A,block_C)$ is a mutex because $block_A$ can only be on top of a single block.

A {\em domain invariant} is an instance-independent invariant, i.e. holds for any possible initial state and set of objects. Therefore, if a given state $s$ holds $s\nvDash \phi$ such that $\phi$ is a {\em domain invariant}, it means that $s$ is not a valid state. Domain invariants are often compactly defined as {\em lifted invariants} (also called schematic invariants)~\cite{rintanen:schematicInvariants:AAAI2017}. For instance, $\phi_2=\forall x:\ (\neg handempty\vee \neg holding(x))$, is a {\em domain mutex} for the {\em blocksworld} because the robot hand is never empty and holding a block at the same time.

\section{Learning \strips\ action models from {\em state-invariants}}
\label{sec:learning}
We define the task of learning a planning action model from {\em state-invariants} as a tuple $\Lambda=\tup{P,\Phi,M}$, where:
\begin{itemize}
\item $P=\tup{F,A[\cdot],I,G}$, is a {\em classical planning problem} where $A[\cdot]$ is a set of actions s.t., the {\em dynamics} of each action $a\in A[\cdot]$ is {\em unknown} (i.e. functions $\rho$ and/or $\theta$ are undefined for $a\in A[\cdot]$).
\item $\Phi$ is a set of {\em state-invariants} that define constraints about the set of possible states.
\item $M$ is the {\em space of possible action models} for the $A[\cdot]$ actions (i.e., the set of possible specifications of the $\rho$ and/or $\theta$ functions for each $a\in A[\cdot]$ action).
\end{itemize}

A model $\mathcal{M}\in M$ is a {\em solution} to the $\Lambda=\tup{P,\Phi,M}$ learning task iff there exists a plan $\pi$ that solves $P=\tup{F,A[\cdot],I,G}$, when the semantics of each action $a\in A[\cdot]$ is given by $\mathcal{M}$, and such that any state traversed by the trajectory $\tau(\pi,P)$ is {\em consistent} with the {\em state-invariants} $\Phi$.

Next, we show that the set $M$ of possible action models can be compactly encoded as a set of propositional variables and a set of constraints over those variables. Then, we show how to exploit this compact encoding to solve a $\Lambda=\tup{P,\Phi,M}$ learning task with an off-the-shelf classical planner.

@@

Despite any element of ${\mathcal I}_{\Psi,\xi}$ can {\em a priori} appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of schema $\xi$, in practice the actual space of possible \strips\ schemata is bounded by constraints of three kinds:
\begin{enumerate}
\item {\bf Syntactic constraints}. \strips\ constraints require $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$. Considering exclusively these syntactic constraints, the size of the space of possible \strips\ schemata is given by $2^{2\times|{\mathcal I}_{\Psi,\xi}|}$. {\em Typing constraints} are also of this kind~\cite{mcdermott1998pddl}. 
\item {\bf Domain-specific constraints}. One can introduce domain-specific knowledge to constrain further the space of possible schemata. For instance, in the {\em blocksworld} one can argue that {\small\tt on($v_1$,$v_1$)} and {\small\tt on($v_2$,$v_2$)} will not appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ lists of an action schema $\xi$ because, in this specific domain, a block cannot be on top of itself. {\it State invariants} are constraints of this kind. 
\item {\bf Observation constraints}. The observation of the actions and states resulting from the execution of a plan depicts {\em semantic knowledge} that constraints further the space of possible action schemata.   
\end{enumerate}

@@

@@

\item Actions for {\em applying} an action model $\xi$ built by the {\em insert} actions and bounded to objects $\omega\subseteq\Omega^{|pars(\xi)|}$ (where $\Omega$ is the set of {\em objects} used to induce the fluents $F$ by assigning objects in $\Omega$ to the $\Psi$ predicates, and $\Omega^k$ is the $k$-th Cartesian power of $\Omega$). The action parameters, $pars(\xi)$, are bound to the objects in $\omega$ that appear in the same position. These actions validate also that any state traversed by $P_{\Lambda}$ solutions is {\em consistent} with the {\em state-invariants} $\Phi$.
\end{enumerate}
\end{itemize}

@@

\subsection{Effective prunning of inconsistent action models with {\em domain mutex}}
We define a {\em domain mutex} as a $(p,q)$ predicates pair where both $p\in\Psi$ and $q\in\Psi$ are predicates that shape the set of fluents $F$ of a given planning problem and such that they satisfy the following formulae $p\leftrightarrow \neg q$ where are the predicate variables are universally quantified. For instance, predicates $holding(x)$ and $clear(x)$ from the {\em blocksworld} are {\em domain mutex} since they satisfy $\forall x\ holding(x)\leftrightarrow\neg clear(x)$ while predicates $clear(x)$ and $ontable(x)$ (also from the {\em blocksworld}) are not {\em domain mutex} because they do not always satisfy $\forall x\ clear(x)\leftrightarrow\neg ontable(x)$.

We pay attention to this particular class of {\em state-invariants} because they define the {\em state-properties} of a given type of objects~\cite{fox:TIM:JAIR1998} and because they enable an effectively pruning of inconsistent /strips/ action models. Our approach to implement this pruning is extending the conditional effects of the $\mathsf{insertPre_{p,\xi}}$ and $\mathsf{insertPre_{p,\xi}}$ actions (i.e., the actions that determine a solution model $\mathcal{M}$) with extra conditional effects indicating that the programmed model is {\em invalid} (i.e., inconsistent with a {\em domain mutex} in $\Phi$). Note that this {\em consistency} checking is more effective than the one implemented at the $\mathsf{apply_{\xi,\omega}}$ actions since $\mathsf{insertPre_{p,\xi}}$ and $\mathsf{insertPre_{p,\xi}}$ actions appear at an earlier stage of the planning process.

@@

\subsection{Learning from partiallly specified models}
The compilation is defined assuming $M$ represents the {\em full} space of \strips\ action models. In some contexts it is however reasonable to assume that the action model is not learned from scratch, e.g. because some parts of the action model are known~\cite{ZhuoNK13,sreedharan2018handling}. Here we show that the compilation approach is also flexible to this particular learning scenario.

When the action model for a \strips\ schema $\xi$ is partially specified, the known preconditions and effects are encoded setting the corresponding fluents $\{pre\_e\_\xi, eff\_e\_\xi\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}$ to true in the initial state. Further, the corresponding insert actions, $\mathsf{insertPre_{p,\xi}}$ and $\mathsf{insertEff_{p,\xi}}$, become unnecessary and are removed from $A_{\Lambda}$, making the classical planning task $P_{\Lambda}$ easier to be solved.

For example, suppose that the preconditions of the {\em blocksworld} action schema {\tt stack} are known, then the initial state $I$ is extended with literals, {\small\tt(pre\_holding\_v1\_stack)} and {\small\tt(pre\_clear\_v2\_stack)} and the associated actions $\mathsf{insertPre_{holding_v1,stack}}$ and $\mathsf{insertPre_{clear_v2,stack}}$ can be safely removed from the $A_{\Lambda}$ action set without altering the {\em soundness} and {\em completeness} of the $P_{\Lambda}$ compilation.

@@

\section{Learning from observations of plan executions}
\label{sec:observations}
Inductive approaches for the learning of planning action models compute an action model starting from an input set of observations of plan executions. Next we formalize the observation model for that kind of knowledge and the we show that the compilation can easily be extended to consider also inductive knowledge in the form of observations of plan executions.

@@

\subsection{Learning from observations with {\em classical planning}}
To consider also inductive knowledge in the form of an $\mathcal{O}(\tau)$ observation of a plan execution, the compilation is extended in the following way:
\begin{itemize}
\item One fluent $\{validated_j\}_{0\leq j\leq m}$ to point at every $s_j^o\in\mathcal{O}(\tau)$ state observation. Two fluents, $at_i$ and $next_{i,i+1}$, {\small $1\leq i \leq n$}, to iterate through the $n$ observed actions of $\mathcal{O}(\tau)$. The former is used to ensure that actions are executed in the same order as they are observed in $\mathcal{O}(\tau)$. The latter is used to iterate to the next planning step when solving $P_{\Lambda}$.
\item Adding $at_1$ and $\{next_{i,i+1}\}$, {\small $1\leq i \leq n$} to the initial state and $validated_m$ to the goals $G$ of the classical planning problem to constrain solution plans to be consistent with all the state observations.
\item Extra conditional effects $\{at_{i},plan(name(a_i),\Omega^{pars(a_i)},i)\}\rhd\{\neg at_{i},at_{i+1}\}_{\forall i\in [1,n]}$ are included in the $\mathsf{apply_{\xi,\omega}}$ actions to ensure that actions are applied in the same order as they appear in $\mathcal{O}(\tau)$.
\item Actions for {\em validating} the partially observed state $s_j^o\in\mathcal{O}(\tau)$, {\tt\small $1\leq j< m$}. These actions are also part of the postfix of the solution plan $\pi_\Lambda$ and they are aimed at checking that the observable data of the input plan trace $\tau$ follows after the execution of the apply actions.
\item One $\mathsf{validate_{j}}$ action to constraint the solution plans to be consistent with the $s_j^o\in\mathcal{O}(\tau)$ input state observation, {\small $(1\leq j\leq m)$}.  
\end{itemize}
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{validate_{j}})=&s_j^o\cup\{validated_{j-1}\},\\
\cond(\mathsf{validate_{j}})=&\{\emptyset\}\rhd\{\neg validated_{j-1}, validated_j\}.
\end{align*}
\end{small}

So far we have explained the extension of the compilation for learning from a single observation $\mathcal{O}(\tau)$. The extension to the more general case of a set of observation $\{\mathcal{O}(\tau_1),\ldots,\mathcal{O}(\tau_k)\}$ is implemented with a small modification. In particular, the actions in $P_{\Lambda}$ for {\em validating} the last state $s_m^o\in \mathcal{O}(\tau_t)$, {\tt\small $1\leq t\leq k$} reset also the current state and the current plan step.

@@

In {\em Inductive Logic Programming} it is very common to make the hypothesis be consistent with some form deductive knowledge apart from the examples, what is ussually called {\em background knowledge}~\cite{muggleton1994inductive}.

>>>>

\begin{abstract}
This paper addresses the learning of action models from {\em state-invariants} (i.e logic formulae that specify constraints about the possible states of a given domain). The benefit of exploiting {\em state-invariants} is two-fold, they allow to reduce the space of possible action models and to complete learning examples that are only partially observed. Our approach for the learning of \strips\ action models from {\em state-invariants} is a {\em classical planning} compilation. The compilation is flexible to different kinds of input knowledge (e.g., partially observations of plan executions including partially observed intermediate states and/or actions) and outputs an action model that is {\em consistent} with the given input knowledge. The experimental results show that, even at unfavorable scenarios where input observations are minimal (just an {\em initial state} and the {\em goals}), {\em state-invariant} are helpful to learn good quality \strips\ action models.
\end{abstract}

@@

The specification of planning action models is a complex process that limits, too often, the application of {\em model-based planning} systems to real-world tasks~\cite{kambhampati:modellite:AAAI2007}. The {\em machine learning} of action models can relieve the {\em knowledge acquisition bottleneck} of planning and nowadays, there exists a wide range of effective approaches for learning action models~\cite{arora:amodels:ker2018}. Many of the most successful approaches for learning planning action models are however purely {\em inductive}~\cite{yang2007learning,pasula2007learning,mourao2010learning,zhuo2013action}, meaning that their performance is linked to the {\em amount} and {\em quality} of the input learning examples.

This paper addresses the learning of action models exploiting a different source of knowledge, {\em deductive} knowledge, to cushion the negative impact of insufficient learning examples. In more detail our approach leverages {\em state-invariants}, i.e. logic formulae that specify constraints about the possible states of a given domain. Given an action model, state-of-the-art planners are used to infer {\em state-invariants} from that model to reduce search spaces and make the planning process more efficient~\cite{helmert2009concise}. In this paper we follow the opposite direction and leverage {\em state-invariants} to learn the planning action model. The benefit of learning \strips\ action models from {\em state-invariants} is two-fold, {\em state-invariants} allow us to reduce the space of possible action models and to complete learning examples that are only partially observed.

Our approach for learning \strips\ action models from {\em state-invariants} is compile the learning task into a classical planning task. Our compilation is flexible to different kinds of input knowledge (e.g., partially/fully observations of actions of plan executions as well as partially/fully observed intermediate states) and outputs an action model that is {\em consistent} with the given input knowledge. The experimental results show that, even at unfavorable scenarios where input observations are minimal (just an {\em initial state} and the {\em goals}), {\em state-invariant} help to learn better \strips\ models than with the existing {\em classical planning} compilation by~\citeauthor{aineto2018learning}~\citeyear{aineto2018learning}.

@@

{\em State invariants} is a kind of state-constraints useful for computing more compact state representations~\cite{helmert2009concise} or making {\em satisfiability planning} and {\em backward search} more efficient~\cite{rintanen2014madagascar,alcazar2015reminder}. Given a classical planning problem $P=\tup{F,A,I,G}$, a {\em state invariant} is a formula $\phi$ that holds at the initial state of a given classical planning problem, $I\models \phi$, and at every state $s$, built from $F$, that is reachable from $I$ by applying actions in $A$. For instance Figure~\ref{fig:strongest-invariant} shows five clauses that define the {\em state invariants} for the {\em blocksworld} planning domain~\cite{slaney2001blocks}. There are infinitely many strongest invariants, but they are all logically equivalent, and computing the strongest invariant is PSPACE-hard (as hard as testing plan existence~\cite{bylander:complexity:AIJ1994}).

\begin{figure}[hbt!]
  \begin{footnotesize}
$\forall x_1,x_2\ ontable(x_1)\leftrightarrow\neg on(x_1,x_2)$.\\
$\forall x_1,x_2\ clear(x_1)\leftrightarrow\neg on(x_2,x_1)$.\\
$\forall x_1,x_2,x_3\ \neg on(x_1,x_2)\vee\neg on(x_1,x_3)\ such\ that\ x_2\neq x_3$.\\
$\forall x_1,x_2,x_3\ \neg on(x_2,x_1)\vee\neg on(x_3,x_1)\ such\ that\ x_2\neq x_3$.\\
$\forall x_1,\ldots,x_n\ \neg(on(x_1,x_2)\wedge on(x_2,x_3)\wedge\ldots\wedge on(x_{n-1},x_n)\wedge on(x_n,x_1)).$
\end{footnotesize}
 \caption{\small Example of {\em state-invariants} for the {\em blocksworld} domain.}
\label{fig:strongest-invariant}
\end{figure}

A {\em mutex} (mutually exclusive) is a state invariant that takes the form of a binary clause and indicates a pair of different properties that cannot be simultaneously true~\cite{kautz:mutex:IJCAI1999}. For instance in a three-block {\em blocksworld}, $\neg on(block_A,block_B)\vee \neg on(block_A,block_C)$ is a {\em mutex} because $block_A$ can only be on top of a single block.

A {\em domain invariant} is an instance-independent invariant, i.e. holds for any possible initial state and set of objects. Therefore, if a given state $s$ holds $s\nvDash \phi$ such that $\phi$ is a {\em domain invariant}, it means that $s$ is not a valid state. Domain invariants are often compactly defined as {\em lifted invariants} (also called schematic invariants)~\cite{rintanen:schematicInvariants:AAAI2017}. For instance, $\forall x:\ (\neg handempty\vee \neg holding(x))$, is a {\em domain mutex} for the {\em blocksworld} because the robot hand is never empty and holding a block at the same time.

\section{Learning \strips\ action models from {\em state-invariants}}
\label{sec:learning}
We define the task of learning a planning action model from {\em state-invariants} as a tuple $\Lambda=\tup{P,\Phi,M}$, where:
\begin{itemize}
\item $P=\tup{F,A[\cdot],I,G}$, is a {\em classical planning problem} where $A[\cdot]$ is a set of actions s.t., the {\em dynamics} of each action $a\in A[\cdot]$ is {\em unknown} (i.e. functions $\rho$ and/or $\theta$ are undefined for $a\in A[\cdot]$).
\item $\Phi$ is a set of {\em state-invariants} that define constraints about the set of possible states in the previous planning problem $P$.
\item $M$ is the {\em space of possible action models} for the $A[\cdot]$ actions (i.e., the set of possible specifications of the $\rho$ and/or $\theta$ functions for each $a\in A[\cdot]$ action).
\end{itemize}

We say that a given model $\mathcal{M}\in M$ is a {\em solution} to the $\Lambda=\tup{P,\Phi,M}$ learning task iff there exists a plan $\pi$ that solves $P=\tup{F,A[\cdot],I,G}$, when the semantics of each action $a\in A[\cdot]$ is given by $\mathcal{M}$, and such that any state traversed by a trajectory $\tau(\pi,P)$ is {\em consistent} with the input set of {\em state-invariants} $\Phi$.

Next, we show that the set $M$, of possible action models, can be compactly encoded as a set of propositional variables and a set of constraints over those variables. Then, we show how to exploit this compact encoding to solve a $\Lambda=\tup{P,\Phi,M}$ learning task with an off-the-shelf classical planner.

@@

Despite any element of ${\mathcal I}_{\Psi,\xi}$ can {\em a priori} appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of schema $\xi$, in practice the actual space of possible \strips\ schemata is bounded by constraints of two kinds:
\begin{enumerate}
\item {\bf Syntactic constraints}. \strips\ constraints require $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$. Considering exclusively these syntactic constraints, the size of the space of possible \strips\ schemata is given by $2^{2\times|{\mathcal I}_{\Psi,\xi}|}$. {\em Typing constraints} are also of this kind~\cite{mcdermott1998pddl}. 
\item {\bf Observation constraints}. The observation of the actions and states resulting from the execution of a plan depicts {\em semantic knowledge} that constraints further the space of possible action schemata.   
\end{enumerate}

@@

In addition, one can introduce {\em domain-specific knowledge} to constrain further the space of possible schemata. For instance, in the {\em blocksworld} one can argue that {\small\tt on($v_1$,$v_1$)} and {\small\tt on($v_2$,$v_2$)} will not appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ lists of an action schema $\xi$ because, in this specific domain, a block cannot be on top of itself. {\it State invariants} are {\em domain-specific knowledge} and they can be seen as either {\em syntactic} or {\em semantic} constraints because on the one hand, they constrain the space of possible action models but on the other hand, they can be used to complete partial observations of the states traversed by a plan.

@@

\item Actions for {\em applying} an action model $\xi$ built by the {\em insert} actions and bounded to objects $\omega\subseteq\Omega^{|pars(\xi)|}$ (where $\Omega$ is the set of {\em objects} used to induce the fluents $F$ by assigning objects in $\Omega$ to the $\Psi$ predicates, and $\Omega^k$ is the $k$-th Cartesian power of $\Omega$). The action parameters, $pars(\xi)$, are bound to the objects in $\omega$ that appear in the same position. These actions validate also that any state traversed by $P_{\Lambda}$ solutions is {\em consistent} with the {\em state-invariants} $\Phi$. The definition $\mathsf{apply_{\xi,\omega}}$ actions is also more compact in our compilation that the one previously proposed by~\citeauthor{aineto2018learning}~\citeyear{aineto2018learning} since are not using disjunctions to code the possible preconditions of an action schema.
\end{enumerate}
\end{itemize}

@@

\subsection{Pruning inconsistent action models with {\em domain mutex}}
We define a {\em domain mutex} as a $(p,q)$ predicates pair where both $p\in\Psi$ and $q\in\Psi$ are predicates that shape the set of fluents $F$ of a given planning problem and such that they satisfy the following formulae $p\leftrightarrow \neg q$ where are the predicate variables are universally quantified. For instance, predicates $holding(x)$ and $clear(x)$ from the {\em blocksworld} are {\em domain mutex} since they satisfy $\forall x\ holding(x)\leftrightarrow\neg clear(x)$ while predicates $clear(x)$ and $ontable(x)$ (also from the {\em blocksworld}) are not {\em domain mutex} because they do not always satisfy $\forall x\ clear(x)\leftrightarrow\neg ontable(x)$.

We pay attention to this particular class of {\em state-invariants} because they define the {\em state-properties} of a given type of objects~\cite{fox:TIM:JAIR1998} and because they enable an effectively pruning of inconsistent \strips\ action models. Our approach to implement this pruning is extending the conditional effects of the $\mathsf{insertPre_{p,\xi}}$ and $\mathsf{insertPre_{p,\xi}}$ actions (i.e., the actions that determine a solution model $\mathcal{M}$) with extra conditional effects indicating that the programmed model is {\em invalid} (i.e., inconsistent with a {\em domain mutex} in $\Phi$). Note that this {\em consistency} checking is more effective than the one implemented at the $\mathsf{apply_{\xi,\omega}}$ actions since $\mathsf{insertPre_{p,\xi}}$ and $\mathsf{insertPre_{p,\xi}}$ actions appear at an earlier stage of the planning process.

@@

@@

\section{Learning from observations of plan executions}
\label{sec:observations}
Inductive approaches for the learning of planning action models compute an action model starting from an input set of observations of plan executions. This section provides a formal model for such input observations and shows how to leverage {\em state-invariants} to automatically {\em complete} those input observations. The section ends with the extension of our compilation to exploit the {\em completed} observations for the learning of \strips\ action models.

@@

\subsection{Completing observations with {\em domain mutex}}
Our observation model follows the {\em open world} assumption, in other words, what is not observed is considered unknown. Here, we show that {\em state-invariants} are helpful to infer new knowledge that was unobserved.

Given a {\em domain mutex} $(p,q)$ and a state observation $s_j^o\in \mathcal{O}(\tau)$, {\small $(1\leq j\leq m)$}, such that the literal $p(\omega)\in s_j^o$ is an instantiation of predicate $p$ over some subset of objects $\omega\subseteq\Omega^{|pars(p)|}$ then, the state observation can be safely completed adding the new literal $\neg q(\omega)$ (despite $\neg q(\omega)$ was actually unobserved). For instance, if the literal {\tt\small holding(blockA)} is observed in a particular blocksword state and we have the {\em domain mutex} $\forall x\ holding(x)\leftrightarrow\neg clear(x)$ in the input set $\Phi$ of {\em state-invariants} we can safely add to the observation the literal {\tt\small $\neg$clear(blockA)} (despite this literal was actually unobserved). The process is repeated for all the observed states in $\mathcal{O}(\tau)$ and all the {\em domain mutex} in $\Phi$ to produce a new completed observation $\mathcal{O}(\tau)'$.

\subsection{Learning from completed observations with {\em classical planning}}
Let be $\mathcal{O}(\tau)'$ an observation completed as explained above, we extend here our compilation to constraint the possible \strips\ models with $\mathcal{O}(\tau)$':
\begin{itemize}
\item One fluent $\{observed_j\}_{0\leq j\leq m}$ to point at every $s_j^o\in\mathcal{O}(\tau)'$ state observation. Two fluents, $at_i$ and $next_{i,i+1}$, {\small $1\leq i \leq n$}, to iterate through the $n$ observed actions in $\mathcal{O}(\tau)'$. The former is used to ensure that actions are executed in the same order as they are observed. The latter is used to iterate to the next planning step when solving $P_{\Lambda}$.
\item Adding $at_1$ and $\{next_{i,i+1}\}$, {\small $1\leq i \leq n$} to the initial state and $observed_m$ to the goals $G$ of the classical planning problem and hence, constrain the solution plans to be consistent with all the state observations.
\item Adding the extra conditional effects {\small $\{at_{i},plan(name(a_i),\Omega^{pars(a_i)},i)\}\rhd\{\neg at_{i},at_{i+1}\}_{\forall i\in [1,n]}$} to the $\mathsf{apply_{\xi,\omega}}$ actions to ensure that actions are applied in the same order as they appear in $\mathcal{O}(\tau)'$.
\item Actions for {\em validating} the partially observed state $s_j^o\in\mathcal{O}(\tau)'$, {\tt\small $1\leq j< m$}. These actions are also part of the postfix of the solution plan $\pi_\Lambda$ and they are aimed at checking that the observable data of the input observation $\mathcal{O}(\tau)$' follows after the execution of the apply actions.
\item One $\mathsf{validate_{j}}$ action to constraint the solution plans to be consistent with the $s_j^o\in\mathcal{O}(\tau)'$ input state observation, {\small $(1\leq j\leq m)$}.  
\end{itemize}
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{validate_{j}})=&s_j^o\cup\{observed_{j-1}\},\\
\cond(\mathsf{validate_{j}})=&\{\emptyset\}\rhd\{\neg observed_{j-1}, observed_j\}.
\end{align*}
\end{small}

So far we explained the extension of the compilation for learning from a single observation $\mathcal{O}(\tau)'$. The extension to the more general case of a set of observation $\{\mathcal{O}(\tau_1),\ldots,\mathcal{O}(\tau_k)\}$ is implemented with a small modification. In particular, the actions in $P_{\Lambda}$ for {\em validating} the last state $s_m^o\in \mathcal{O}(\tau_t)$, {\tt\small $1\leq t\leq k$} reset also the current state and the current plan step.

@@

In {\em Inductive Logic Programming} it is very common to make the hypothesis be consistent with some form deductive knowledge apart from the examples, what is usually called {\em background knowledge}~\cite{muggleton1994inductive}.
